# Paired t-test

This is a test used to analyze the difference between the means of two treatments/groups where the samples are paired. The two treatments are applied to separate, independent samples.

We have a dataset with the weights of 17 girls before and after treatment for anorexia. Since the same individuals were measured twice, we can use a paired t-test to see if the treatment worked.

```{r}
#| include: FALSE
library(tidyverse)
anorexia <- read_csv('datasets/anorexia.csv')
```

Before we start, we can store the differences in each pair as another variable and tidy the data, converting to long format. I'm also storing the row numbers as a column so that we know which before row corresponds to which after row, and ordering them such that Prior shows up before Post (as opposed to alphabetical order).

```{r}
anorexia <- anorexia %>%
  mutate(diff = Post - Prior, id = row_number()) %>%
  pivot_longer(c('Prior', 'Post'), names_to = 'time', values_to = 'weight')

anorexia$time <- ordered(anorexia$time, levels=c('Prior', 'Post'))
```

## Visualizing data

A typical way to visualize the differences for a paired t-test is using a plot like this. Note that we will have to pivot the data to long format.

```{r}
#| fig-align: center
anorexia %>%
  ggplot(aes(x = time, y = weight)) +
  geom_point(shape = 1) +
  geom_line(aes(group = id), color = 'grey') +
  xlab('Time') +
  ylab('Weight (lbs)')
```

## Hypotheses

The hypothesis of a paired t-test is all about the mean of the difference in the pairs, $\mu_d$:

* $H_0$: The mean change in weight after anorexia treatment was 0.
* $H_A$: The mean change in weight after anorexia treatment was not 0.

Or symbolically, it can be represented as:

* $H_0: \mu_d = 0$
* $H_A: \mu_d \neq 0$

## Checking assumptions

The assumptions for conducting a paired t-test are as follows:

* The individuals are randomly sampled from the population.
* The differences are normally distributed in the population.

Since the sample size is so small, a normal quantile plot may not be as helpful in evaluating the normality of the data. Alternatively, it can be verified by conducting a Shapiro-Wilk normality test, which checks the null hypothesis that the data are sampled from a normal distribution:

```{r}
shapiro.test(anorexia$diff)
```

Since the P-value is greater than 0.05, we fail to reject the $H_0$ of normality and can proceed with the test.

## Performing a paired t-test

We can use the built-in `t.test` method in R to perform this test. If you need help using this command, you can use `?t.test` to view the [documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) for this command.

```{r}
t.test(weight ~ time, data = anorexia,
  paired = TRUE, conf.level = 0.95)
```

Since the test produced a P-value less than 0.05 ($\alpha$), we can reject the null hypothesis.
