# Analysis of Variance

An ANOVA test with two groups is equivalent to a two-sample t-test. Unlike two-sample t-tests, ANOVA can simultaneously test many groups without inflating the Type I error rate beyond $\alpha$.

Here is an example of using ANOVA to test whether the mean flipper length is different between three different species of penguins.

```{r}
#| include: FALSE
library(tidyverse)
palmerpenguins <- read_csv('../datasets/palmerpenguins.csv') 
```

## Visualizing data

Since we have large sample sizes, it would be helpful to visualize the data with a violin plot (or a boxplot would work too).

```{r}
#| fig-align: center
palmerpenguins %>%
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_violin() +
  geom_boxplot(width = 0.1) +
  geom_jitter(size = 1, shape = 1, width = 0.15) +
  xlab('Species') +
  ylab('Flipper length (mm)')
```

## Hypotheses

The null hypothesis for an ANOVA test is that the population means, $\mu_i$, are the same for all treatments. The alternate hypothesis is that at least one population mean is different from the rest. ANOVA cannot give us information about _which_ group's population mean is different from the rest.

* $H_0$: Mean flipper length is equal among all three species ($\mu_1 = \mu_2 = \mu_3$).
* $H_A$: At least one species' mean flipper length is different from the rest.

We'll be conducting a fixed-effects ANOVA test, which uses the F-test statistic.

## Checking assumptions

ANOVA makes the same assumptions as a two-sample t-test:

* Both samples represent random samples obtained from their populations.
* The variable in question is normally distributed in each population.
* The variance of the variable is the same in each population.

We can check the normality assumption using a normal quantile plot:

```{r}
#| fig-align: center
palmerpenguins %>%
  ggplot(aes(sample = flipper_length_mm)) +
  stat_qq(shape = 1) +
  stat_qq_line() +
  facet_grid(~ species) +
  xlab('Normal quantile') +
  ylab('Flipper length (mm)')
```

We can check the homoscedasticity (equal variances) assumption using the Bartlett's test:

```{r}
bartlett.test(flipper_length_mm ~ species, data = palmerpenguins)
```

Since the P-value is greater than 0.05, we fail to reject the null hypothesis of equal variances.

## Performing an ANOVA test

There are two steps to performing an ANOVA test in R. First, we need to create a linear model using the `lm` function. Then, we can pass that to the `anova` method to generate the appropriate ANOVA results.

```{r}
palmerpenguins.lm <- lm(flipper_length_mm ~ species, data = palmerpenguins)
anova(palmerpenguins.lm)
```

||$df$|Sum of Squares|Mean of Squares|$F$-ratio|$P$|
|-|-|-|-|-|-|
|Species|2|50526|25262.9|567.41|2.2e-16|
|Residuals|330|14693|44.5|||

Due to the low P-value, we can reject the null hypothesis that the mean flipper length is equal. But now the question is, which group mean(s) differ?

## Post-hoc Tukey

The Tukey-Kramer post-hoc test (yes, a mouthful) compares all group means to all other group means. This is preferable to doing multiple two-sample t-tests because the latter inflates the Type I error rate.

To conduct this test in R, we first need to re-do the ANOVA test using the `aov` method and pass that onto `TukeyHSD()`. An example is shown below:

```{r}
aov(flipper_length_mm ~ species, data = palmerpenguins) %>%
  TukeyHSD(conf.level = 0.95)
```
